---
title: "COVID-19 Data by State"
author: "kyoungwan.kang@cgu.edu"
date: "4/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```


```{r import-packages}
library(tidyverse)
library(lubridate)
```


```{r import-data}

rawdata <- read.csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")

poprawdata <- read.csv("https://www2.census.gov/programs-surveys/popest/datasets/2010-2020/national/totals/nst-est2020.csv")

# SUMLEV Geographic summary level 
# REGION Census Region code 
# DIVISION Census Division code 
# STATE State FIPS code 
# NAME State name
# POPESTIMATE2020

maskflagdf <- read.csv("covid-19_mask_by_state.csv")

```

# COVID-19 Historical U.S. State-Level Data from the New York Times
Coronavirus (Covid-19) Data in the United States: [https://github.com/nytimes/covid-19-data](https://github.com/nytimes/covid-19-data)
Datafile: U.S. State-Level Data ([Raw CSV](https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv))

U.S. tracking page by the New York Times - [Link](https://www.nytimes.com/interactive/2020/us/coronavirus-us-cases.html)

```{r}
tail(rawdata, 10)
```



# 2020 State Population Estimates
<Annual Estimates of the Resident Population for the United States, States, and the District of Columbia: April 1, 2010 to July 1, 2020 (NST-EST2020)> 

[State Population Totals: 2010-2019](https://www.census.gov/data/datasets/time-series/demo/popest/2010s-state-total.html)

Source: U.S. Census Bureau, Population Division
Dataset file [link](https://www2.census.gov/programs-surveys/popest/datasets/2010-2020/nation/NST-EST2020.csv)

```{r}
tail(poprawdata, 10)
```

# States with mask mandates

* [These States Have COVID-19 Mask Mandates (04/12/2021)](https://www.usnews.com/news/best-states/articles/these-are-the-states-with-mask-mandates) from [U.S.News](https://www.usnews.com)
  + States with mask mandates (27 states + D.C.)
  California, Colorado, Connecticut, Delaware, District of Columbia, Hawaii, Illinois, Kentucky, Louisiana,
  Maine, Maryland, Massachusetts, Michigan, Minnesota, Nevada, New Hampshire, New Jersey, New Mexico, New York,
  North Carolina, Ohio, Oregon, Pennsylvania, Rhode Island, Vermont, Virginia, Washington, West Virginia 
  
  + States without mask mandates (23 states)
  Alabama, Alaska, Arizona, Arkansas, Florida, Georgia, Idaho, Indiana, Iowa, Kansas, Mississippi, Missouri,
  Montana, Nebraska, North Dakota, Oklahoma, South Carolina, South Dakota, Tennessee, Texas, Utah, Wisconsin,
  Wyoming

* [First Trust COVID Tracker 03/05/21](https://www.ftportfolios.com/Blogs/EconBlog/2021/3/5/coronavirus-high-frequency-data-030521) - [one page report](https://www.ftportfolios.com/common/contentfileloader.aspx?contentguid=a942b521-8099-40ed-aaa2-203fd59bc396)
  + States with mask mandates
  + States without mask mandates (16 states)
  Alaska, Arizona, Florida, Georgia, Idaho, Iowa, Mississippi, Missouri, Montana, Nebraska, North Dakota,
  Oklahoma, South Carolina, South Dakota, Tennessee, Texas


States showing up in the map in the one page report being used for this analysis

```{r}
# Get data on the latest day
thelatestdate <- max(rawdata[[1]])

# populations for states
popdf <-
  poprawdata %>%
  select (state = NAME, pop = POPESTIMATE2019) 

# Create a data frame, casedf, by using spread function (cf. reshape::cast)
casedf <-
  rawdata %>%
  select(state, fips, date, cases) %>%
  spread(key = date,value = cases)

# Get numbers of columns in casedf
numcol <- ncol(casedf)

# Replace NA with 0
i <- 0
for (i in 3:numcol) {
  casedf[[i]] <- replace_na(casedf[[i]],0)
}

# Copy casedf to dailycasedf to calculate daily new case
dailycasedf <- casedf

# Calculate daily new cases and put them to each column of date in dailycasedf
for (i in 4:numcol) {
  dailycasedf[[i]] = casedf[[i]] - casedf[[(i-1)]]
}

# Copy state population in popdf and state names & fips in casedf to dailycapitadf 
# No populations in 
dailycapitadf <- left_join(popdf[6:56,], dailycasedf, by = "state")

# Calculate daily cases per 100K people
for (i in 4:ncol(dailycapitadf)) {
  dailycapitadf[[i]] = round(((dailycapitadf[[i]] / dailycapitadf[[2]]) * 100000), digits = 2)
}
```

```{r}
# Calculate weekly average cases per 100K people
numcol_dcapita <- ncol(dailycapitadf)
  
# Change format to long with pivot_longer() (cf. reshape::melt(), dplyr::gather())
tempcapita <-
  dailycapitadf %>%
  pivot_longer(
    cols = colnames(dailycapitadf)[4:numcol_dcapita],
    names_to = "date",
    values_to = "casepercapita"
  )

# Add week column by using lubridate::year() and lubridate::week() (year-week format)
tempcapita <- 
  tempcapita %>%
  mutate(
    week = paste0(epiyear(date), "-", if_else(nchar(x <- epiweek(date)) > 1, as.character(x), paste0("0", x)))
  )

# Create weeklycapita data frame and calculate weekly average cases per 100K by state
weeklycapita <-
  tempcapita %>% group_by(state, week) %>%
  summarize(weeklymean = round(mean(casepercapita), digits = 2))

# Subset between 12/01/2020 and 03/02/2021 
startweek <- paste0("2020-", as.character(epiweek("2020-12-01")))
lastweek <- paste0("2021-", if_else(nchar(x <- epiweek("2021-03-02")) > 1, as.character(x), paste0("0", x)))

ourdata <-
weeklycapita %>%
  filter(week >= startweek & week <= lastweek)
```


```{r}
# Add mask flag to the data frame
ourdata <- ourdata %>%
  spread(key = week, value = weeklymean)

ourdata <- 
  left_join(
    ourdata,
    select(maskflagdf, state, flag = flag.0302),
    by = "state"
  ) %>%
  select (
    state,
    colnames(.)[16],
    colnames(.)[2:15]
  ) %>%
  pivot_longer(
    .,
   cols = colnames(.)[3:16],
   names_to = "week",
   values_to = "weeklymean"
  )

```

# Weekly average cases per 100K people by state

```{r}
ggplot(
  data = ourdata,
  mapping = aes(
    x = week,
    y = weeklymean
  )    
) + 
  geom_point(
    mapping = aes(
    group = state,
    color = flag
    )  
  ) +
  geom_smooth(
    mapping = aes(
    group = flag,
    color = flag
    )  
  ) +
  scale_x_discrete(
    breaks = c("2020-49", "2021-01", "2021-05", "2021-09")
  )
  
```

* Red dots: States without mask mandates
* Blue dots: States with mask mandates
* Red and blue lines: LOESS (locally estimated scatterplott smoothing): a generalization of moving average and polynomial regression. - [Wikipedia link](https://en.wikipedia.org/wiki/Local_regression)
* X axis
  + 2020-49: 49th week of 2020, the first day: Sunday, `r as.character(cut(ymd("2020-01-01") + weeks(48), "weeks", start.on.monday = F))`
  + 2021-01: 1st week of 2021, the first day: Sunday, `r as.character(cut(ymd("2021-01-01") + weeks(0), "weeks", start.on.monday = F))`
  + 2021-05: 5th week of 2021, the first day: Sunday, `r as.character(cut(ymd("2021-01-01") + weeks(0), "weeks", start.on.monday = F))``r as.character(cut(ymd("2021-01-01") + weeks(4), "weeks", start.on.monday = F))`
  + 2021-09: 9th week of 2021, the first day: Sunday, `r as.character(cut(ymd("2021-01-01") + weeks(8), "weeks", start.on.monday = F))`
  


```{r}

# create a summary data frame
tmpdf <- ourdata

# Change week to the first day of the week
# as.character(cut(ymd("2020-01-01") + weeks(48) - 1, "weeks", start.on.monday = F))

# Calculate weekly average for states with and without mask mandates
sumdf <- tmpdf %>%
  group_by(flag, week) %>%
  summarize(
    weeklyavg = round(mean(weeklymean), digits = 2)
  ) %>% ungroup()
```


# Weekly average cases per 100K people by states with mask mandaes and ones without mask mandaes
```{r}
# Plot
ggplot(
  data = sumdf,
  mapping = 
      aes(
        x = week,
        y = weeklyavg,
        group = flag,
        color = flag
      )
  ) +
  geom_line() +
  geom_smooth(method = loess) +
  scale_x_discrete(
    breaks = c("2020-49", "2021-01", "2021-05", "2021-09")
  )

```

* Blue lines: states with mask mandates (bold line show up LOESS )
* Red lines: states without mask mandates (bold line show up LOESS)
* X axis
  + 2020-49: 49th week of 2020, the first day: Sunday, `r as.character(cut(ymd("2020-01-01") + weeks(48), "weeks", start.on.monday = F))`
  + 2021-01: 1st week of 2021, the first day: Sunday, `r as.character(cut(ymd("2021-01-01") + weeks(0), "weeks", start.on.monday = F))`
  + 2021-05: 5th week of 2021, the first day: Sunday, `r as.character(cut(ymd("2021-01-01") + weeks(0), "weeks", start.on.monday = F))``r as.character(cut(ymd("2021-01-01") + weeks(4), "weeks", start.on.monday = F))`
  + 2021-09: 9th week of 2021, the first day: Sunday, `r as.character(cut(ymd("2021-01-01") + weeks(8), "weeks", start.on.monday = F))`

# Weekly average cases per 100K people in states with mask mandaes
```{r}
# Plot
ggplot(
  data = filter(sumdf, flag == "Y"),
  mapping = 
      aes(
        x = week,
        y = weeklyavg,
        group = flag,
        color = flag
      )
  ) +
  geom_line(color = "blue") +
  geom_smooth(method = loess) +
  scale_x_discrete(
    breaks = c("2020-49", "2021-01", "2021-05", "2021-09")
  ) + 
  ylim(0, 100)
```

* Blue line: weekly average cases per 100K people in the states 
* Red line: LOESS
* Grey area: Standard Error 
* X axis
  + 2020-49: 49th week of 2020, the first day: Sunday, `r as.character(cut(ymd("2020-01-01") + weeks(48), "weeks", start.on.monday = F))`
  + 2021-01: 1st week of 2021, the first day: Sunday, `r as.character(cut(ymd("2021-01-01") + weeks(0), "weeks", start.on.monday = F))`
  + 2021-05: 5th week of 2021, the first day: Sunday, `r as.character(cut(ymd("2021-01-01") + weeks(0), "weeks", start.on.monday = F))``r as.character(cut(ymd("2021-01-01") + weeks(4), "weeks", start.on.monday = F))`
  + 2021-09: 9th week of 2021, the first day: Sunday, `r as.character(cut(ymd("2021-01-01") + weeks(8), "weeks", start.on.monday = F))`


# Weekly average cases per 100K people in states without mask mandaes
```{r}
# Plot
ggplot(
  data = filter(sumdf, flag == "N"),
  mapping = 
      aes(
        x = week,
        y = weeklyavg,
        group = flag,
        color = flag
      )
  ) +
  geom_line(color = "blue") +
  geom_smooth(method = loess) +
  scale_x_discrete(
    breaks = c("2020-49", "2021-01", "2021-05", "2021-09")
  ) +
  ylim(0, 100)
```

* Blue line: weekly average cases per 100K people in the states 
* Red line: LOESS
* Grey area: Standard Error
* X axis
  + 2020-49: 49th week of 2020, the first day: Sunday, `r as.character(cut(ymd("2020-01-01") + weeks(48), "weeks", start.on.monday = F))`
  + 2021-01: 1st week of 2021, the first day: Sunday, `r as.character(cut(ymd("2021-01-01") + weeks(0), "weeks", start.on.monday = F))`
  + 2021-05: 5th week of 2021, the first day: Sunday, `r as.character(cut(ymd("2021-01-01") + weeks(0), "weeks", start.on.monday = F))``r as.character(cut(ymd("2021-01-01") + weeks(4), "weeks", start.on.monday = F))`
  + 2021-09: 9th week of 2021, the first day: Sunday, `r as.character(cut(ymd("2021-01-01") + weeks(8), "weeks", start.on.monday = F))`